Detection of Ecstatic Interactions in High-Dimensional Genomic Data Using Random Forests

### 2025-05-04 epistatic_interactions.tex
changed names to epistatic_interactions.tex

\usepackage{threeparttable}
\usepackage{url}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[figuresleft]{rotating}

%usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}

@@

\author[1,2,3]{Hawlader A. Al-Mamun\corref{cor1}}
\ead{halmamun@intergrain.com}
\author[4]{Rob Dunne} %% Author name
\author[5]{Ross L. Tellam}
\author[6]{Klara Verbyla}

@@

\affiliation[2]{organization={InterGrain Pty Ltd.},%Department and Organization
  addressline={19 Ambitious Link}, 
  city={Bibra Lake},
  postcode={6163}, 
  state={WA},
  country={Australia}}

\affiliation[3]{organization={School of Biological Sciences},%Department and Organization
  addressline={The University of Western Australia}, 
  city={Crawley},
  postcode={6009}, 
  state={WA},
  country={Australia}}

\affiliation[4]{organization={Data61, CSIRO},%Department and Organization
  addressline={13 Garden Street}, 
  city={Eveleigh},
  postcode={2015}, 
  state={NSW},
  country={Australia}}

@@

\affiliation[6]{organization={The Center for Aquaculture Technologies},%Department and Organization
  addressline={8445 Camino Santa Fe, Suite 104}, 
  city={San Diego},
  postcode={92121}, 
  state={California},
  country={USA}}

@@

There are several options described in  \cite{Gauran.et.al.2018}.
Using the ``one stage procedure'' and $C_1$ constant (see supplementary file Supplementary_File_Rheumatoid_Data.Rmd)
we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.

@@

\subsection{threshholding the data}
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this in number of ways  (breakpoints in the distribution,
  entropy tests and others. See the supplementary information file Supplementary_File_Rheumatoid_Data.Rmd) and then take the median. In
  the case of the Rheumatoid data, the selected breakpoint was 2 which gave us 222 interactions;
\item test each interaction with a linear model as described below.
\end{itemize}

@@

\begin{sidewaystable}
  \tiny
 \begin{center}
   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
     \hline
  SNP pair & D'         & r$^2$ & SNP1 location & \makecell{SNP1 odds\\ ratio} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} &
  \makecell{SNP1 \\$p$-value$^d$} & SNP2 location & \makecell{SNP2 odds\\ ratio} & \makecell{SNP2 odds \\ ratio (95\% CI)$^b$ } &
  \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction\\ Odds ratio} &\makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} \\
   rs3134926:rs910050     & 0.10 & 0.01 & 6:32232370  & 0.61 & 0.55,0.6  & 1.99E-20 & 6:32347877 & 0.79 & 0.72,0.86 &  7.11E-05& 1.77 & 1.53,2.06 & 3.38E-11 \\
 rs3134926:rs4530903      & 0.14 & 0.01 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32614112 & 1.41 & 1.24,1.59 &  6.59E-05& 1.64 & 1.34,2.01 & 9.90E-04 \\
 rs9431614:rs9296009      & 0.01 & 0.00 & 1:230944623 & 1.00 & 0.89,1.11 & 5.17E+02 & 6:32146738 & 1.80 & 1.64,1.97 &  2.51E-33& 1.62 & 1.35,1.96 & 2.91E-04 \\
 rs3134926:rs9391858      & 0.06 & 0.00 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32373621 & 1.24 & 1.11,1.39 &  5.76E-02& 1.55 & 1.28,1.88 & 4.22E-03 \\
 rs10456058:rs9273363$^a$ & 0.08 & 0.00 & 6:31526961  & 1.11 & 1.03,1.21 & 5.97E+00 & 6:32658495 & 1.34 & 1.23,1.46 &  3.39E-08& 1.50 & 1.32,1.7  & 2.73E-07 \\
 rs910050:rs6457617       & 0.17 & 0.02 & 6:32347877  & 0.79 & 0.72,0.86 & 7.11E-05 & 6:32696074 & 0.44 & 0.4,0.48  &  1.06E-69& 1.39 & 1.22,1.59 & 4.32E-04 \\
 rs9268831:rs615672       & 0.15 & 0.01 & 6:32459971  & 0.69 & 0.63,0.75 & 3.37E-16 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 1.34 & 1.17,1.53 & 9.12E-03 \\
 rs2076533:rs1270689      & 0.03 & 0.00 & 6:32395750  & 0.49 & 0.45,0.54 & 4.60E-55 & 7:12961515 & 1.06 & 0.98,1.16 &  8.23E+01& 1.31 & 1.16,1.48 & 1.48E-02 \\
 rs17421624:rs615672      & 0.18 & 0.01 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 0.73 & 0.64,0.84 & 7.58E-03 \\
 rs6457617:rs2621382$^e$      & 0.17 & 0.03 & 6:32696074  & 0.44 & 0.4,0.48  & 1.06E-69 & 6:32792668 & 1.02 & 0.94,1.11 &  3.44E+02& 0.73 & 0.64,0.83 & 6.46E-04 \\
 rs17421624:rs3134926     & 0.08 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.72 & 0.62,0.83 & 7.12E-03 \\
 rs17421624:rs427037      & 0.03 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:3224448  & 0.70 & 0.62,0.78 &  2.70E-07& 0.69 & 0.59,0.82 & 1.11E-0  \\
rs9296009:rs4530903       & 0.10 & 0.00 & 6:32146738  & 1.80 & 1.64,1.97 & 2.51E-33 & 6:3261411  & 1.41 & 1.24,1.59 &  6.59E-05& 0.67 & 0.55,0.81 & 3.16E-0  \\
rs206015:rs3134926        & 0.06 & 0.00 & 6:32214982  & 1.49 & 1.33,1.67 & 2.90E-09 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.67 & 0.55,0.82 & 4.66E-0  \\
\hline
\end{tabularx}
% \end{tabular}
 \end{center}
 \caption{Eight pairs of interacting SNPs have significant odds ratio after pairwise LD evaluation}
 \label{table:eight.pairs}
 \begin{tablenotes}%
 \item[$^{a}$] rs10456058 has merged into rs2734573.
 \item[$^{b}$]  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP.
 \item[$^{c}$]  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
 \item[$^{d}$]  Bonferroni adjusted p-value.
 \item[$^e$]   The pair of SNPs used as an example in section \label{section:A_linear_model_example}.
 \end{tablenotes}
\end{sidewaystable}

@@

Key points:
\begin{itemize}
\item Detecting epistatic interactions from high dimensional genomic data is very challenging due to the computational
  burden induced by the large number of required combinatorial test.
\item Using a Random Forests algorithm, we provided a computationally simple way to detect pairwise interactions.
\item The method produced reliable results in the case of simulated and interpretable results in several real-world
  data sets
\end{itemize}

@@

\bibliographystyle{elsarticle-num}
\bibliography{./manuscript.bib}

>>>>

\usepackage{threeparttable}
\usepackage{url}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{placeins}
%\usepackage{tikz}
%\usetikzlibrary{calc}
\usepackage[figuresleft]{rotating}
\usepackage{longtable, lscape, booktabs, array}
%usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}

@@

\author[1]{Hawlader A. Al-Mamun\corref{cor1}\fnref{fn1}}
\ead{halmamun@intergrain.com}
\author[4]{Rob Dunne} %% Author name
\author[5]{Ross L. Tellam}
\author[1]{Klara Verbyla, \fnref{fn2}}

@@

% \affiliation{organization={InterGrain Pty Ltd.},
%   addressline={19 Ambitious Link}, 
%   city={Bibra Lake},
%   postcode={6163}, 
%   state={WA},
%   country={Australia}}
% and
% \affiliation{organization={School of Biological Sciences},%Department and Organization
%   addressline={The University of Western Australia}, 
%   city={Crawley},
%   postcode={6009}, 
%   state={WA},
%   country={Australia}}

\fntext[fn1]{Currently at InterGrain Pty Ltd, 19 Ambitious Link, Bibra Lake, 6163, WA and
  School of Biological Sciences, The University of Western Australia, Crawley, 6009, WA Australia}

\affiliation[4]{organization={Data61, CSIRO},
  addressline={13 Garden Street}, 
  city={Eveleigh},
  postcode={2015}, 
  state={NSW},
  country={Australia}}

@@

% \affiliation[6]{organization={The Center for Aquaculture Technologies},%Department and Organization
%   addressline={8445 Camino Santa Fe, Suite 104}, 
%   city={San Diego},
%   postcode={92121}, 
%   state={California},
%   country={USA}}
\fntext[fn2]{Currently at The Center for Aquaculture Technologies, 8445 Camino Santa Fe, Suite 104, San Diego, 92121,
  California, USA}

%%Research highlights
\begin{highlights}
\item The method detects epistatic interactions in high dimensional genomic data.
\item Computationally efective and produces reliable results in the case of simulated and interpretable results in several real-world
  data sets
\end{highlights}

@@

There are several options described in  \cite{Gauran.et.al.2018}.
Using the ``one stage procedure'' and $C_1$ constant (see supplementary file Supplementary\_File\_Rheumatoid\_Data.Rmd)
we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.

@@

\subsection{threshholding the data}
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this in number of ways  (breakpoints in the distribution,
  entropy tests and others. See the supplementary information file Supplementary\_File\_Rheumatoid\_Data.Rmd) and then take the median. In
  the case of the Rheumatoid data, the selected breakpoint was 2 which gave us 222 interactions;
\item test each interaction with a linear model as described below.
\end{itemize}

@@

\begin{sidewaystable}
  \tiny
 \begin{center}
   \begin{tabularx}{\textwidth}{|c|c|c|p{1.16cm}|p{0.8cm}|p{1.16cm}|c|c|p{0.8cm} |c|c|c|c|p{1cm}||}
 \hline
 SNP pair & D' & r$^2$ & 
      \rotatebox{60}{SNP1 location} & 
      \rotatebox{60}{\makecell{SNP1 odds\\ ratio}} & 
      \rotatebox{60}{\makecell{SNP1 odds ratio \\(95\% CI)\textsuperscript{b}}} &
      \rotatebox{60}{SNP1 $p$-value\textsuperscript{d}} & 
      \rotatebox{60}{SNP2 location} & 
      \rotatebox{60}{\makecell{SNP2 odds\\ ratio}} & 
      \rotatebox{60}{\makecell{SNP2 odds ratio \\ (95\% CI)\textsuperscript{b}}} &
      \rotatebox{60}{SNP2 $p$-value\textsuperscript{d}} & 
      \rotatebox{60}{\makecell{Interaction \\Odds ratio}} &
      \rotatebox{60}{\makecell{Interaction Odds ratio\\ (95\% CI)\textsuperscript{c}}} & 
      \rotatebox{60}{\makecell{Interaction\\ $p$-value\textsuperscript{d}}} \\
      \hline
   rs3134926:rs910050     & 0.10 & 0.01 & 6:32232370  & 0.61 & 0.55,0.6  & 1.99E-20 & 6:32347877 & 0.79 & 0.72,0.86 &  7.11E-05& 1.77 & 1.53,2.06 & 3.38E-11 \\
 rs3134926:rs4530903      & 0.14 & 0.01 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32614112 & 1.41 & 1.24,1.59 &  6.59E-05& 1.64 & 1.34,2.01 & 9.90E-04 \\
 rs9431614:rs9296009      & 0.01 & 0.00 & 1:230944623 & 1.00 & 0.89,1.11 & 5.17E+02 & 6:32146738 & 1.80 & 1.64,1.97 &  2.51E-33& 1.62 & 1.35,1.96 & 2.91E-04 \\
 rs3134926:rs9391858      & 0.06 & 0.00 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32373621 & 1.24 & 1.11,1.39 &  5.76E-02& 1.55 & 1.28,1.88 & 4.22E-03 \\
 rs10456058:rs9273363$^a$ & 0.08 & 0.00 & 6:31526961  & 1.11 & 1.03,1.21 & 5.97   & 6:32658495 & 1.34 & 1.23,1.46 &  3.39E-08& 1.50 & 1.32,1.7  & 2.73E-07 \\
 rs910050:rs6457617       & 0.17 & 0.02 & 6:32347877  & 0.79 & 0.72,0.86 & 7.11E-05 & 6:32696074 & 0.44 & 0.4,0.48  &  1.06E-69& 1.39 & 1.22,1.59 & 4.32E-04 \\
 rs9268831:rs615672       & 0.15 & 0.01 & 6:32459971  & 0.69 & 0.63,0.75 & 3.37E-16 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 1.34 & 1.17,1.53 & 9.12E-03 \\
 rs2076533:rs1270689      & 0.03 & 0.00 & 6:32395750  & 0.49 & 0.45,0.54 & 4.60E-55 & 7:12961515 & 1.06 & 0.98,1.16 &  8.23E+1& 1.31 & 1.16,1.48 & 1.48E-02 \\
 rs17421624:rs615672      & 0.18 & 0.01 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 0.73 & 0.64,0.84 & 7.58E-03 \\
 rs6457617:rs2621382$^e$      & 0.17 & 0.03 & 6:32696074  & 0.44 & 0.4,0.48  & 1.06E-69 & 6:32792668 & 1.02 & 0.94,1.11 &  3.44E+02& 0.73 & 0.64,0.83 & 6.46E-04 \\
 rs17421624:rs3134926     & 0.08 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.72 & 0.62,0.83 & 7.12E-03 \\
 rs17421624:rs427037      & 0.03 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:3224448  & 0.70 & 0.62,0.78 &  2.70E-07& 0.69 & 0.59,0.82 & 1.11E-03  \\
rs9296009:rs4530903       & 0.10 & 0.00 & 6:32146738  & 1.80 & 1.64,1.97 & 2.51E-33 & 6:3261411  & 1.41 & 1.24,1.59 &  6.59E-05& 0.67 & 0.55,0.81 & 3.16E-03  \\
rs206015:rs3134926        & 0.06 & 0.00 & 6:32214982  & 1.49 & 1.33,1.67 & 2.90E-09 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.67 & 0.55,0.82 & 4.66E-03  \\
\hline
\end{tabularx}
% \end{tabular}
 \end{center}
 \caption{Eight pairs of interacting SNPs have significant odds ratio after pairwise LD evaluation}
 \label{table:eight.pairs}
 \begin{tablenotes}%
 \item $^{a}$ rs10456058 has merged into rs2734573.
 \item $^{b}$  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP.
 \item $^{c}$  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
 \item $^{d}$  Bonferroni adjusted p-value.
 \item $^e$   The pair of SNPs used as an example in section \label{section:A_linear_model_example}.
 \end{tablenotes}
\end{sidewaystable}

@@

Key points:
\begin{itemize}
\item Detecting epistatic interactions from high dimensional genomic data is very challenging due to the computational
  burden induced by the large number of required combinatorial test.
\item Using a Random Forests algorithm, we provided a computationally simple way to detect pairwise interactions.
\item The method produced reliable results in the case of simulated and interpretable results in several real-world
  data sets.
\end{itemize}

@@

\bibliographystyle{elsarticle-num}
\bibliography{./epistatic_interactions.bib}
